Estimating duration in the presence of misclassification

### 2024-07-29 MCMC-appendix.tex
Rob typo comments

A common way of constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution is using the Metropolis--Hastings algorithm~\autocite{hastingsMCMC}.
The algorithm requires a proposal distribution $q(\psi' \mid \psi)$, which proposes that the chain moves to a new state $\psi'$ from its current state $\psi$; $\pi(\psi)$, a calculation of $p(\psi \mid y)$ up to some constant.
At each iteration, a proposal is draw from $q$ and then accepted or rejected based on the \emph{acceptance ratio}, computed using $\pi$.
The full algorithm is given in \cref{inc-prev:algorithm:MH}, with further details in available in a wide-range of sources~\autocites[e.g.][]{brooksMCMCNotes}[chapter 11]{gelmanBDA}.
\begin{algorithm}
 set $\psi^0$ to some initial value\;
 \For{$i = 1, \dots, M$}{
  sample a proposal $\psi' \sim q(\psi' \mid \psi)$ \;
  calculate the acceptance probability $\alpha(\psi, \psi') = \min \left( 1, \frac{\pi(\psi') q(\psi \mid \psi')}{\pi(\psi) q(\psi' \mid \psi)} \right)$ \;
  with probability $\alpha(\psi, \psi')$ set $\psi^i$ to $\psi'$, otherwise set $\psi^i$ to $\psi^{i-1}$ \;
 }
 \caption{The Metropolis--Hastings algorithm.}
 \label{inc-prev:algorithm:MH}
\end{algorithm}

@@

Various metrics quantify this, I make use of the improved Rhat statistic~\autocite{vehtariRhat}.
A Rhat of 1 means that the within-chain and between-chain variance are equal, informally that convergence is perfect, as far as can be measured by this technique.
Various rules-of-thumb exist for what should be considered an acceptable Rhat.
Historically, 1.1 has commonly been recommended, although more recent work suggests that 1.01 or 1.05 is more appropriate~\autocite{vehtariRhat}.

>>>>

A common way of constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution is using the Metropolis--Hastings algorithm~\autocite{hastingsMCMC}.
The algorithm requires a proposal distribution $q(\psi' \mid \psi)$, which proposes that the chain moves to a new state $\psi'$ from its current state $\psi$; $\pi(\psi)$, a calculation of $p(\psi \mid y)$ up to some constant.
At each iteration, a proposal is draw from $q$ and then accepted or rejected based on the \emph{acceptance ratio}, computed using $\pi$.
The full algorithm is given in \cref{inc-prev:algorithm:MH}, with further details in available in a wide-range of sources~\autocites[e.g.][]{brooksMCMCNotes}[chapter 11]{gelmanBDA}.
\begin{algorithm}
 set $\psi^0$ to some initial value\;
 \For{$i = 1, \dots, M$}{
  sample a proposal $\psi' \sim q(\psi' \mid \psi^{i-1})$ \;
  calculate the acceptance probability $\alpha(\psi, \psi') = \min \left( 1, \frac{\pi(\psi') q(\psi \mid \psi')}{\pi(\psi) q(\psi' \mid \psi)} \right)$ \;
  with probability $\alpha(\psi, \psi')$ set $\psi^i$ to $\psi'$, otherwise set $\psi^i$ to $\psi^{i-1}$ \;
 }
 \caption{The Metropolis--Hastings algorithm.}
 \label{inc-prev:algorithm:MH}
\end{algorithm}

@@

Various metrics quantify this; I make use of the improved Rhat statistic~\autocite{vehtariRhat}.
A Rhat of 1 means that the within-chain and between-chain variance are equal, informally that convergence is perfect, as far as can be measured by this technique.
Various rules-of-thumb exist for what should be considered an acceptable Rhat.
Historically, 1.1 has commonly been recommended, although more recent work suggests that 1.01 or 1.05 is more appropriate~\autocite{vehtariRhat}.

### 2024-07-29 cis-imperfect-testing.tex
Rob p110 comment
Rob typo comments

In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of detectability was developed.
However, the initial application of this framework produced an implausibly short estimate.
In this chapter, I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
I then propose several models of false negatives (\cref{imperf-test:sec:simulate}) and extend the model of \cref{E-perf-test} to incorporate one of them (\cref{imperf-test:sec:modelling})
After showing through simulation that the new model recovers the true duration distribution (\cref{imperf-test:sec:sim-study-results}), I apply it to the CIS data (\cref{imperf-test:sec:application}).
Finally, I discuss the results and further work in \cref{imperf-test:sec:discussion}.

@@

$\vec{\nu}'_k \in \set{E}'$ occurs if and only if all the following conditions are met.
\begin{enumerate}
  \item $\vec{\nu}_k \in \set{E}$.
  \item The elements of $\vec{y}_k$ corresponding to the tests at times $r_k^{(b)}$ and $l_k^{(e)}$ are positive, \ie $y_{k,1} = y_{k,m_k-1} = 1$.
  \item The element of $\vec{y}_k$ corresponding to the test at time $r_k^{(e)} + 1$ is negative, \ie $y_{k,m_k} = 0$.
\end{enumerate}
The final two conditions are due to the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.

>>>>

In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of detectability was developed.
However, the initial application of this framework produced an implausibly short estimate.
In this chapter, I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
I then propose several models of false negatives (\cref{imperf-test:sec:simulate}) and extend the model of \cref{E-perf-test} to incorporate one of them (\cref{imperf-test:sec:modelling}).
After showing through simulation that the new model recovers the true duration distribution (\cref{imperf-test:sec:sim-study-results}), I apply it to the CIS data (\cref{imperf-test:sec:application}).
Finally, I discuss the results and further work in \cref{imperf-test:sec:discussion}.

@@

The elements of $\set{E}'$ are all $\vec{\nu}_k'$ such that:
\begin{enumerate}
  \item $\vec{\nu}_k \in \set{E}$.
  \item The elements of $\vec{y}_k$ corresponding to the tests at times $r_k^{(b)}$ and $l_k^{(e)}$ are positive, \ie $y_{k,1} = y_{k,m_k-1} = 1$.
  \item The element of $\vec{y}_k$ corresponding to the test at time $r_k^{(e)} + 1$ is negative, \ie $y_{k,m_k} = 0$.
\end{enumerate}
The final two conditions are due to the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.

### 2024-07-29 cis-perfect-testing.tex
Rob p101 comment
Rob p102 comment
Rob typo comments

The final component of \cref{perf-test:eq:full-posterior} required is $1 - p_u$.
Recall the definition $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$.
Therefore:
\begin{align}
  1 - p_u
  &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset, i(j) = i \mid \vec{\theta}) \\
  &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}) P(i(j) = i \mid \vec{\theta}) \\
  &= 1 - \frac{1}{\Ncis}\sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}) \\
  &= \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} (1 - \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}))
  \label{perf-test:eq:pu}
\end{align}
assuming $P(i(j) = i \mid \vec{\theta}) = 1/\Ncis$ as before.
Let $p_{iu} = \prob(O(j) = \emptyset \mid i(j) = i, \vec{\theta})$.
Therefore, the crucial component is $1 - p_{iu}$.
This is one minus the probability that an episode in individual $i$ was undetected, \ie the probability of the episode being detected.

@@

\subsection{Setup}

I simulate a dataset of detected episodes with the same characteristics as that in the CIS by the following procedure.
\begin{enumerate}
    \item Extract the test schedules for each individual who had at least one test during the period of interest.
    \item Draw an episode start time, $b_{j}$ for each individual uniformly at random between 2 July 2020 ($t = -100$) and 6 December 2020 ($t = T$).
    \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_{j} = b_{j} + d_i - 1$.
    \item Simulate the test results based on the test schedule, $b_{j}$, and $e_{j}$. Tests between $b_{j}$ and $e_{j}$ (inclusive) are positive, all other tests are negative.
    \item Discard episodes where there are no positive tests (\ie undetected episodes) and then apply the same inclusions/exclusions as in \cref{perf-test:sec:problem}.
    \item Of these remaining episodes, sample \numprint{4800} to match the sample size of the true dataset. This is needed because in step 2 the entire cohort was infected, while in the real study only an (unknown) portion is infected.
    \item For this final set of episodes, calculate $(l_j^{(b)}, r_j^{b}, l_j^{(e)}, r_j^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
\end{enumerate}

@@

The prior on the total number of infections was centred on the true value with a dispersion parameter of 1.

@@

I implemented the inference in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
Convergence was assessed using Rhat and ESS as described in \cref{E-MCMC:sec:convergence}, and all runs checked for divergent transitions.
This meant that the prior was very diffuse.

@@

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/survival-results}
  \caption[Comparison of survival function estimates under different priors]{Comparison of the posterior estimate of the survival distribution (pointwise median and 95\% CrI) when using different priors on the hazard. Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown. y-axis has either natural (left) or log (right) scale. I show only the first 100 days of the distribution for readability. \label{perf-test:fig:survival-results}}
\end{figure}

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/hazard-results}
  \caption[Comparison of hazard estimates under different priors]{Comparison of the posterior estimate of the hazard (pointwise median and 95\% CrI) when using different priors on the hazard. Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown. \label{perf-test:fig:hazard-results}}
\end{figure}

>>>>

The final component of \cref{perf-test:eq:full-posterior} required is $1 - p_u$.
Recall the definition $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$.
Therefore:
\begin{align}
  1 - p_u
  &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset, i(j) = i \mid \vec{\theta}) \\
  &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i, \vec{\theta}) P(i(j) = i \mid \vec{\theta}) \\
  &= 1 - \frac{1}{\Ncis}\sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i, \vec{\theta}) \\
  &= \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} (1 - \prob(O(j) = \emptyset \mid i(j) = i, \vec{\theta}))
  \label{perf-test:eq:pu}
\end{align}
assuming $P(i(j) = i \mid \vec{\theta}) = 1/\Ncis$ as before.
Let $p_{iu} = \prob(O(j) = \emptyset \mid i(j) = i, \vec{\theta})$.
Therefore, the crucial component is $1 - p_{iu}$.
This is one minus the probability that an episode in individual $i$ was undetected, \ie the probability of the episode being detected.

@@

\subsection{Setup} \label{perf-test:sec:sim-setup}

I simulate a dataset of detected episodes with the same characteristics as that in the CIS by the following procedure.
\begin{enumerate}
    \item Extract the test schedules for each individual who had at least one test during the period of interest.
    \item Draw an episode start time, $b_{j}$ for each individual uniformly at random between 2 July 2020 ($t = -100$) and 6 December 2020 ($t = T$).
    \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_{j} = b_{j} + d_i - 1$.
    \item Simulate the test results based on the test schedule, $b_{j}$, and $e_{j}$. Tests between $b_{j}$ and $e_{j}$ (inclusive) are positive, all other tests are negative.
    \item Discard episodes where there are no positive tests (\ie undetected episodes) and then apply the same inclusions/exclusions as in \cref{perf-test:sec:problem}.
    \item Of these remaining episodes, sample \numprint{4800} to match the sample size of the true dataset. This is needed because in step 2 the entire cohort was infected, while in the real study only an (unknown) portion is infected.
    \item For this final set of episodes, calculate $(l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
\end{enumerate}

@@

The prior on the total number of infections was centred on the true value with a dispersion parameter of 1.
This meant that the prior was very diffuse.

@@

I implemented the inference in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
Convergence was assessed using Rhat and ESS as described in \cref{E-MCMC:sec:convergence}, and all runs checked for divergent transitions.

@@

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/survival-results}
  \caption[Comparison of survival function estimates under different priors]{Comparison of the posterior estimate of the survival distribution (pointwise median and 95\% CrI) when using different priors on the hazard. Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown. y-axis has either natural (left) or log (right) scale. I show only the first 100 days of the distribution for readability. Black line shows the value used in the simulation (see \cref{perf-test:sec:sim-setup}). \label{perf-test:fig:survival-results}}
\end{figure}

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/hazard-results}
  \caption[Comparison of hazard estimates under different priors]{Comparison of the posterior estimate of the hazard (pointwise median and 95\% CrI) when using different priors on the hazard. Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown. Black line shows the value used in the simulation (see \cref{perf-test:sec:sim-setup}). \label{perf-test:fig:hazard-results}}
\end{figure}

### 2024-07-29 distributions.tex
Rob typo comments

\begin{landscape}
\section{Continuous distributions}
\begin{tabular}{llp{3.5cm}lll}
Name & Notation & Parameters & Mean & Variance & pdf\\
\hline \\
Normal & $\N(\mu, \sigma^2)$ & $\mu$ (mean)\newline $\sigma^2 > 0$ (variance) & $\mu$ & $\sigma^2$ & $\frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{(x-\mu)^2}{2\sigma^2}}$ \\
\makecell{Multivariate \\ normal} & $\MNorm(\vec{\mu}, \matr{\Sigma})$ & $\mu \in \reals^k$ (mean vector)\newline $\Sigma \in \reals^{k \times k}$ (covariance matrix, positive semi-definite) & $\vec{\mu}$ & $\matr{\Sigma}$ & $(2 \pi)^{-k / 2} \operatorname{det}(\matr{\Sigma})^{-1 / 2} \exp \left(-\frac{1}{2}(\vec{x}-\vec{\mu})^T \matr{\Sigma}^{-1}(\vec{x}-\vec{\mu})\right)$ \\
Lognormal & $\LN(\mu, \sigma^2)$ & $\mu$ (mean of the logarithm)\newline $\sigma^2 > 0$ (variance of the logarithm) & $e^{\mu + \frac{\sigma^2}{2}}$ & $(e^{\sigma^2} - 1)e^{2\mu + \sigma^2}$ & $\frac{1}{x\sigma\sqrt{2\pi}}e^{-\frac{(\ln(x)-\mu)^2}{2\sigma^2}}$ \\
Gamma & $\GamDist(\alpha, \beta)$ & $\alpha > 0$ (shape) \newline $\beta > 0$ (rate) & $\frac{\alpha}{\beta}$ & $\frac{\alpha}{\beta^2}$ & $\frac{\beta^\alpha}{\Gamma(\alpha)\theta}x^{\alpha-1}e^{-\beta x}$ \\
Beta & $\BetaDist(\alpha, \beta)$ & $\alpha > 0$ (shape)\newline $\beta > 0$ (shape) & $\frac{\alpha}{\alpha+\beta}$ & $\frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$ & $\frac{1}{B(\alpha,\beta)}x^{\alpha-1}(1-x)^{\beta-1}$ \\
Exponential & $\Exponential(\lambda)$ & $\lambda$ (rate) & $\frac{1}{\lambda}$ & $\frac{1}{\lambda^2}$ & $\lambda e^{-\lambda x}$ \\
LKJ* & $\LKJ(\eta)$ & $\eta > 0$ (shape) & $\matr{I}$ (identity matrix) & N/A & $C \det(\matr{x})^{\eta-1}$ \\
\end{tabular}

>>>>

\begin{landscape}
\section{Continuous distributions}
\begin{tabular}{llp{3.5cm}lll}
Name & Notation & Parameters & Mean & Variance & pdf\\
\hline \\
Normal & $\N(\mu, \sigma^2)$ & $\mu$ (mean)\newline $\sigma^2 > 0$ (variance) & $\mu$ & $\sigma^2$ & $\frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{(x-\mu)^2}{2\sigma^2}}$ \\
\makecell{Multivariate \\ normal} & $\MNorm(\vec{\mu}, \matr{\Sigma})$ & $\mu \in \reals^k$ (mean vector)\newline $\Sigma \in \reals^{k \times k}$ (covariance matrix, positive semi-definite) & $\vec{\mu}$ & $\matr{\Sigma}$ & $(2 \pi)^{-k / 2} \operatorname{det}(\matr{\Sigma})^{-1 / 2} \exp \left(-\frac{1}{2}(\vec{x}-\vec{\mu})^T \matr{\Sigma}^{-1}(\vec{x}-\vec{\mu})\right)$ \\
Lognormal & $\LN(\mu, \sigma^2)$ & $\mu$ (mean of the logarithm)\newline $\sigma^2 > 0$ (variance of the logarithm) & $e^{\mu + \frac{\sigma^2}{2}}$ & $(e^{\sigma^2} - 1)e^{2\mu + \sigma^2}$ & $\frac{1}{x\sigma\sqrt{2\pi}}e^{-\frac{(\ln(x)-\mu)^2}{2\sigma^2}}$ \\
Gamma & $\GamDist(\alpha, \beta)$ & $\alpha > 0$ (shape) \newline $\beta > 0$ (rate) & $\frac{\alpha}{\beta}$ & $\frac{\alpha}{\beta^2}$ & $\frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-\beta x}$ \\
Beta & $\BetaDist(\alpha, \beta)$ & $\alpha > 0$ (shape)\newline $\beta > 0$ (shape) & $\frac{\alpha}{\alpha+\beta}$ & $\frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$ & $\frac{1}{B(\alpha,\beta)}x^{\alpha-1}(1-x)^{\beta-1}$ \\
Exponential & $\Exponential(\lambda)$ & $\lambda$ (rate) & $\frac{1}{\lambda}$ & $\frac{1}{\lambda^2}$ & $\lambda e^{-\lambda x}$ \\
LKJ* & $\LKJ(\eta)$ & $\eta > 0$ (shape) & $\matr{I}$ (identity matrix) & N/A & $C \det(\matr{x})^{\eta-1}$ \\
\end{tabular}

### 2024-07-29 incidence-prevalence.tex
Rob typo comments

\emph{Duration}, or \emph{time-to-event}, distributions can be expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
The duration distribution I will be concerned with is the distribution of the duration of detectability.
Consider an individual $i$ infected on day $B_i$ that is prevalent during the inclusive interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}); if $B_i = E_i$ then $i$ is prevalent for exactly one day during which they are detectable.
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$; therefore, with a duration $D_i = E_i - B_i + 1$.
I model the $D_i$s as a discrete iid (independent and identically distributed) random variables; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
The distribution of $D_i$ can be expressed through the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda_{d} = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
Throughout this thesis, the survival function will be the primary quantity of interest, and normally parameterized in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda_{i})$.

@@

Considering multiple values of $t$ makes \cref{inc-prev:eq:EPt} into a system of simultaneous equations.
Therefore, the system can be written in matrix form:
\begin{align}
  \E(\vec{P} \mid \vec{Z}) = \matr{S} \vec{Z}. \label{inc-prev:eq:EPt-matrix} \\
\intertext{$\matr{S}$ is a lower-triangular matrix formed from $S$:}
    (\matr{S})_{i,j} &= \begin{cases}
        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
        0 & \text{otherwise.}
    \end{cases} \label{inc-prev:eq:Smatrix}
\end{align}
where $\vec{P} = [P_1, P_2, \dots]$.
In \cref{E-backcalc}, I will need to solve for this system for $\vec{Z}$ given a known $\hat{P} = \E(\vec{P} \mid {Z})$.
Assuming $\vec{P}$ and $\vec{Z}$ have the same dimension then $\matr{S}$ is lower-triangular with strictly positive diagonal elements; therefore, $\matr{S}$ is invertible, and the system has a unique solution~\autocite[424]{bernsteinFacts}.
The lower-triangular structure of $\matr{S}$ makes allows solving the system through forward-substitution, an efficient method that avoids explicit matrix inversion~\autocite{cormenMatrix}.
Forward-substitution sequentially solves the $t$th element of $\vec{Z}$ ($t = 1, \dots, T$) by calculating:
\begin{align}
Z_t
&= \frac{\hat{P}_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
&= \hat{P}_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
\label{inc-prev:eq:forward-substitute}
\end{align}
where $S(t)$ for $t > \dmax$ is defined to be 0.

>>>>

\emph{Duration}, or \emph{time-to-event}, distributions can be expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
The duration distribution I will be concerned with is the distribution of the duration of detectability.
Consider an individual $i$ infected on day $B_i$ that is prevalent during the inclusive interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}); if $B_i = E_i$ then $i$ is prevalent for exactly one day during which they are detectable.
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$; therefore, with a duration $D_i = E_i - B_i + 1$.
I model the $D_i$s as a discrete iid (independent and identically distributed) random variables; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
The distribution of $D_i$ can be expressed through the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(d - 1)$ with $S(1) = 1$.
%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda_{d} = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
Throughout this thesis, the survival function will be the primary quantity of interest, and normally parameterized in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda_{i})$.

@@

Considering multiple values of $t$ makes \cref{inc-prev:eq:EPt} into a system of simultaneous equations.
Therefore, the system can be written in matrix form:
\begin{align}
  \E(\vec{P} \mid \vec{Z}) = \matr{S} \vec{Z}. \label{inc-prev:eq:EPt-matrix} \\
\intertext{$\matr{S}$ is a lower-triangular matrix formed from $S$:}
    (\matr{S})_{i,j} &= \begin{cases}
        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
        0 & \text{otherwise.}
    \end{cases} \label{inc-prev:eq:Smatrix}
\end{align}
where $\vec{P} = [P_1, P_2, \dots]$.
In \cref{E-backcalc}, I will need to solve for this system for $\vec{Z}$ given a known $\hat{P} = \E(\vec{P} \mid {Z})$.
Assuming $\vec{P}$ and $\vec{Z}$ have the same dimension then $\matr{S}$ is lower-triangular with strictly positive diagonal elements; therefore, $\matr{S}$ is invertible, and the system has a unique solution~\autocite[424]{bernsteinFacts}.
The lower-triangular structure of $\matr{S}$ allows solving the system through forward-substitution, an efficient method that avoids explicit matrix inversion~\autocite{cormenMatrix}.
Forward-substitution sequentially solves the $t$th element of $\vec{Z}$ ($t = 1, \dots, T$) by calculating:
\begin{align}
Z_t
&= \frac{\hat{P}_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
&= \hat{P}_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
\label{inc-prev:eq:forward-substitute}
\end{align}
where $S(t)$ for $t > \dmax$ is defined to be 0.

### 2024-07-29 introduction.tex
Rob p24 comment
Rob p25 comment
Rob typo comments

Incidence-type data originate from observations of the consequences of an infection, \eg the development of symptoms or admission to hospital.
This type of data typically suffers from reporting biases, most problematically that the proportion of infections detected varying over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.

@@

Data on the severe consequences of an infection typically less affected by biases.
An extreme example is the development of AIDS following infection by HIV.
Here, the disease is so severe that it is extremely unlikely for an individual to develop the disease without seeking healthcare~\autocite{evansCompleteness}.

@@

Next, are semi-mechanistic models.
These relate incidence in the past to current (and future) incidence but do not explicitly represent the susceptible population.
A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic,fraserRenewal}.
In discrete time, it takes the form:
\begin{align}
    \E(Z_t \mid Z_0, \dots, Z_{t-1}) &= \sum_{\tau=1}^t \beta(t,\tau) Z_{t-\tau} d \tau
    \label{intro:eq:renewal}
\end{align}
where $Z_t$ is the expected incidence at time $t$ and $\beta(t, \tau)$ is the expected rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
The model parameters can be used to define an effective reproduction number as either $\Re = \sum_{\tau=0}^\infty \beta(t, \tau)$ or $\Re = \sum_{\tau=0}^\infty \beta(t+\tau, \tau)$~\autocite{gosticPractical}.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{thompsonImproved,gosticPractical}.

@@

Solving \cref{intro:eq:disease-model} for $Z$ is an ill-posed problem.
The forwards error of the solution is large when $\max_{i > 0} f(i) / f(0)$~\autocite[chapter 8.2]{highamAccuracy}.
A large forwards error means that the computed value is far from the true solution~\autocite[6--9]{highamAccuracy}; this is avoided by either using an approach which gives a large $f(0)$ or smoothing $Z_t$.

@@

\Textcite{abbottCISincidence} applied the backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
They used summary statistics of prevalence estimates from the CIS, taken from the ONS website, as data; specifically, the prevalence's posterior point estimate and bounds of the 95\% credible interval.
From these summary statistics, they derive a normally-distributed likelihood, independent for each day.
They used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing as $f$ in their observation model; these estimates are based on a longitudinal study in healthcare workers (see \cref{intro:sec:previous-duration-estimates}).
% They fit to summary statistics of the CIS data, using a daily, independent normal likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
\Textcite{colmanAscertainment} used a similar method to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
% Their methodology was similar to \textcite{abbottCISincidence}, including using estimates of $f$ from \textcite{hellewellPCRSensitivity}.
% Rather than using a smooth prior to provide stability, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} assume the ascertainment rate is piecewise constant.
% They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.

>>>>

Incidence-type data originate from observations of the consequences of an infection, \eg the development of symptoms or admission to hospital.
This type of data typically suffers from reporting biases, most problematically that the proportion of infections detected varies over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.

@@

Data on the severe consequences of an infection are typically less affected by biases.
An extreme example is the development of AIDS following infection by HIV.
Here, the disease is so severe that it is extremely unlikely for an individual to develop the disease without seeking healthcare~\autocite{evansCompleteness}.

@@

Next, are semi-mechanistic models.
These relate incidence in the past to current (and future) incidence but do not explicitly represent the susceptible population.
A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic,fraserRenewal}.
In discrete time, it takes the form:
\begin{align}
    \E(Z_t \mid Z_0, \dots, Z_{t-1}) &= \sum_{\tau=1}^t \beta(t,\tau) Z_{t-\tau}
    \label{intro:eq:renewal}
\end{align}
where $Z_t$ is the expected incidence at time $t$ and $\beta(t, \tau)$ is the expected rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
The model parameters can be used to define an effective reproduction number as either $\Re = \sum_{\tau=0}^\infty \beta(t, \tau)$ or $\Re = \sum_{\tau=0}^\infty \beta(t+\tau, \tau)$~\autocite{gosticPractical}.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{thompsonImproved,gosticPractical}.

@@

Solving \cref{intro:eq:disease-model} for $Z$ is an ill-posed problem.
The forwards error of the solution is large when $\max_{i > 0} f(i) / f(0)$~\autocite[chapter 8.2]{highamAccuracy} is large.
A large forwards error means that the computed value is far from the true solution~\autocite[6--9]{highamAccuracy}; this is avoided by either using an approach which gives a large $f(0)$ or by smoothing $Z_t$.

@@

\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
They used summary statistics of prevalence estimates from the CIS, taken from the ONS website, as data; specifically, the prevalence's posterior point estimate and bounds of the 95\% credible interval.
From these summary statistics, they derive a normally-distributed likelihood, independent for each day.
They used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing as $f$ in their observation model; these estimates are based on a longitudinal study in healthcare workers (see \cref{intro:sec:previous-duration-estimates}).
% They fit to summary statistics of the CIS data, using a daily, independent normal likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
\Textcite{colmanAscertainment} used a similar method to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
% Their methodology was similar to \textcite{abbottCISincidence}, including using estimates of $f$ from \textcite{hellewellPCRSensitivity}.
% Rather than using a smooth prior to provide stability, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} assume the ascertainment rate is piecewise constant.
% They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.
